\mytitle

### 2020-05-14 paper.tex
Edits for Section 4 and 5. Comments left for Arik.

In this section, we 
present an empirical evaluation of the two quadratic models based on 
12 standard WFLO instances.
Our main results are: \begin{itemize}
	\item WFL-QUBO when solved using the digital annealer (DA) achieves a similar performance to state-of-the-art exact methods.
		\item The runtime for solving WFL-QUBO on the DA is at least two orders of magnitude lower than WFL-QCOP for hard WFLO instances.  
	\item Given sufficient runtime, solving the WFL-QCOP model yields superior results on most of the instances.
\end{itemize}
We start by outlining the experimental design of our evaluation, followed by an overview of our main results,
and conclude the section with a discussion of the results and their practical implications.

@@

To test and benchmark our models, we used 12 standard benchmark WFLO instances
from the literature~\cite{MOSETTI1994105,Zhang2014,grady2005placement} that we denote with the set $\mathcal{W}$. 
The 12 benchmarks result from varying 2 wind settings (1 and 36 wind directions),
2 grid sizes ($10\times10$ with cell size $200$ meters and $20\times20$ with cell size $100$ meters),
and 3 variations for number of turbines ($m \in \{20, 30, 40\}$). 
For the $36$ wind directions setting, the velocities came from the probability distribution presented in Figure~\ref{fig:prob_wind}.
%We denote $\mathcal{W}$ the set of 12 instances.
Instances with $36$ wind regimes are considered computationally more challenging~\cite{Zhang2014}.

Additional parameters that were used to calculate 
the velocities ($u_{id\infty}, u_{ijd}$) include: turbine height ($60$ meters),
ground roughness ($0.3$ meters), turbine rotor radius ($20$ meters), and a wind farm of size $2km \times 2km$\footnote{The implementation of constructing the 12 standard WFLO instances can be found in~\url{https://bit.ly/2YIjTl1}.}.

@@

The existing constraint programming approach was
reported to perform worse than the ILP model~\cite{Zhang2014} and so was omitted from our the experiments.
We used state-of-the-art commercial software, Gurobi~\cite{gurobi}, to solve WFL-QCOP, ILP-SS and QP-SS. 
For consistency, we shall refer to WFL-QCOP problem as QP-LS as it is a quadratic program that uses the LS objective.
In addition, we 
used both Fujitsu's Digital Annealer and Gurobi to solve WFL-QUBO: we shall refer to the former as DA-LS, and to the latter
as QUBO-LS.

@@

For the DA, we used a single configuration that was found to work best across the different instances. 
The DA was run for increasingly large time limits ranging from $0.7$ seconds to $65$ seconds where we no longer observe an improvement in 
performance. The penalty terms $\lambda_1$ and $\lambda_2$ (see~Eq.~(\ref{QUBO}) are selected such that $\lambda_1= \lambda_2$ and sum of the two penalty terms, namely $\lambda_1+\lambda_2$, 
is set to the highest value that the Digital Annealer allows\footnote{The DA has an upper limit on the size of objective function coefficients.}.

@@

We measured the total expected energy of a solution using the SS expression (see Eq.~(\ref{eq:ss})).
%\footnote{The total mean energy is presented in kW.}.
The only method that directly optimizes SS is QP-SS. Hence, for the other methods, we converted solutions
LS-based solution into the SS expression using Eq.~(\ref{eq:ss}) as proposed in Zhang et al.\ \cite{Zhang2014}. Specifically, let $x = (x_1, \ldots, x_k), k=n^2$ be a turbine placement solution obtained by an LS-based method.
Then, the SS based energy expression can be written as, 
$$\sum_{i\in \mathcal{N}}\sum_{d \in \mathcal{D}} \frac{1}{3} x_i \Bigg(u_{id,\infty} \Bigg[1-\sqrt{\sum_{j\in \mathcal{U}_{id}}x_j\bigg( 1-\frac{u_{ijd}}{u_{id,\infty}} \bigg) } \Bigg]    \Bigg)^3.$$

Next, we tested the performance of our approach as function of solution time. We 
compared the average performance of all WFLO methods on all 12 WFLO instances. Since energy levels are not directly comparable between the different instances (due to varying number of turbines and wind farm specifications) we used the \emph{mean relative error} (MRE) measure. The MRE is defined with respect to the best solution obtained per instance and can be computed over time. Then, the
time-varying MRE can be averaged across the 12 scenarios and presented over time. In what follows, we demonstrate the construction of the MRE measure.

Recall that $\mathcal{W}$ is the set of our 12 standard WFLO instances. 
Let $b_w$ be the best energy solution (in terms of SS) obtained for an instance $w \in \mathcal{W}$ 
across the different methods.
%For example, the best solution (within 3600 seconds) for the second instance $w_2$ that aims at placing $30$ turbines on a $10\times10$ grid with a single wind regime is attained by QUBO-LS and the SS energy expression is $15742$ (kW); thus, we set $b_w = 15742$.
Then, for every point in time $t \in \{0,\ldots, 3600\}$ (measured in seconds), we can compute the relative error of a given approach per instance $w$. Let $\mathcal{A}$ be the set of solution approaches, $\mathcal{A}= \{QUBO-LS, QP-LS, DA-LS, ILP-LS, QP-SS\}$. Further, denote by $b_{w,t,a}$ the best solution attained before time $t$ of approach $a$ for instance $w$. Then, the mean relative error at time $t$ for approach $a$ on instance $w$ is given by \begin{equation}MRE(w,t,a) = \frac{b_w - b_{w,t,a}}{b_w}.\end{equation} Note that the expression is always positive, since we are solving a maximization problem.

We can now average over the different instances at time $t$. The average performance of approach $a$ at time $t$, $MRE(t,a)$ can be computed as, \begin{equation} MRE(t,a) = \frac{1}{|\mathcal{W}|}\sum_{ w\in\mathcal{W}} MRE(w,t,a).\end{equation}
We present $MRE(t,a)$ for each of the approaches to demonstrate the speed of convergence of the proposed LS-based methods.

@@

We observe that the DA-based solution provides
comparable results to state-of-the-art exact methods. For the harder instances (of $36$ wind directions),
the three quadratic LS-based approaches (DA-LS, QUBO-LS, and QP-LS) dominate the others by finding the best (or equal) solutions in 11 out of 12 instances. For easier instances, 
the linear approach (ILP-LS) finds better solutions.
%\todo{@Chris, Jason: do we have a good explanation for this?}. NO
Table~\ref{tab:results2} shows the same type of results as in Table~\ref{tab:results1}, but now after $3600$ seconds. For the exact technique, we can see that the improvement between 100 and 3600 seconds of run-time is marginal. We also observe that the DA is, therefore, still competitive with the other approaches (recall, its solution was found after $10$ seconds),
while QP-LS finds the best solutions in 10 out of the 12 scenarios. 
Our quadratic approaches dominate all six hard scenarios (12 wind directions). %\todo{@Chris: I am not sure that the first table contributes anything beyond the second table. JCB: ``We include the second table to demonstrate that only minor improvements are achieved by the exact techniques between 100 seconds of run-time.''}

@@

The results show that on average (across the 12 instances),
the ILP-LS method starts with poor solutions, yet ends up improving over QP-SS and QUBO-LS after around 1250 seconds.

@@

To demonstrate the speedup gained from using the DA, we compare the point in time when the DA reaches 
a competitive solution (10 seconds) to the point in time when QP-LS reaches the same solution in terms of energy.
Table~\ref{speedup} summarizes the performance of QP-LS when compared to that of the DA after 10 seconds.
We see a speedup factor of X. \todo{discuss with Chris}
%This is an encouraging result, especially due to the pessimistic computational complexity of WFLO.

@@

Our result show empirically that the quadratic programming models that we
introduced in Section~\ref{sec:QUBO4WFLO} (WFL-QCOP and WFL-QUBO) are useful in two ways.
First, the two problems are the basis for the QP-LS and QUBO-LS methods that solved with Gurobi. The two methods, namely QP-LS and QUBO-LS, turn out to be powerful approaches to solving WFLO. Specifically, QP-LS yields the best performance (after 3600 seconds) for most problem instances
and especially,
for the hard instances with 36 wind directions. In addition, the runtime performance of QP-LS is second
only to the Digital Annealer.  Second, the QUBO model can be solved with recent specialized hardware, Fujitsu's DA, as demonstrated by the DA-LS approach. DA-LS yields the best results in 7 out of 12 
instances, shows competitive performance on the other instances, and outperforms all methods in terms of runtime (\todo{speed-up ratio of ...})

@@

@@

In this work, we presented a novel quadratic programming framework
for solving 
wind farm layout optimization. 
The framework is embodied in two models: constrained (WFL-QCOP) and unconstrained binary
models (WFL-QUBO). The former can be used to solve WFLO using advanced optimization software,
while the latter can be mounted onto quantum-inspired machines that can quickly (approximately) solve WFLO. We used the constrained quadratic formulation to prove 
that the computational complexity of WFLO is $\mathcal{NP}$-hard. However,
our empirical evaluation shows that when exact solvers like Gurobi~\cite{gurobi} are
coupled with our WFL-QCOP and WFL-QUBO models they are able to solve small WFLO instances to optimality in less than 3600 seconds.
We also show that when an advanced hardware is used with the QUBO representation, large WFLO instances can be solved quickly with no significant degradation in the total energy compared to exact methods.
This runtime improvement and state-of-the-art performance in terms of the total energy of the resulting solutions makes our quadratic framework useful for designing new wind farms, while varying multiple input parameters.

>>>>

In this section, we 
present an empirical evaluation of the two quadratic models based on 
12 standard WFLO instances.
Our main results are: \begin{itemize}
	\item WFL-QUBO when solved using the digital annealer (DA) achieves a solution quality as good or better than the state-of-the-art exact methods on 10 of 12 instances.
		\item To achieve such a solution quality, the runtime for solving WFL-QUBO on the DA is at least two orders of magnitude lower than WFL-QCOP for hard WFLO instances.  
	\item Given sufficient runtime, solving the WFL-QCOP model yields superior results on most of the instances. \todo{Arik: Is this correct? Table 2 shows DA@10seconds with the best solution found for 7/12 instances and QP-LS@3600 for 6/12. And 4/6 vs. 3/6 for the WR36). Are you comparing QP-LS with QP-SS and ILP-LS? We seem to be underplaying DA performance in this comment.}  
\end{itemize}

We start by outlining the experimental design of our evaluation, followed by an overview of our main results,
and conclude the section with a discussion of the results and their practical implications.

@@

To test and benchmark our models, we used 12 standard benchmark WFLO instances
from the literature~\cite{MOSETTI1994105,Zhang2014,grady2005placement} that we denote with the set $\mathcal{W}$. 
The 12 benchmarks result from varying 2 wind settings (1 and 36 wind directions),
2 grid sizes ($10\times10$ with cell size $200$ meters and $20\times20$ with cell size $100$ meters),
and 3 variations for number of turbines ($m \in \{20, 30, 40\}$). 
For the $36$ wind directions setting, the velocities come from the probability distribution presented in Figure~\ref{fig:prob_wind}.
%We denote $\mathcal{W}$ the set of 12 instances.
Instances with $36$ wind regimes are considered computationally more challenging~\cite{Zhang2014}.

Additional parameters that were used to calculate 
the velocities ($u_{id\infty}, u_{ijd}$) include: turbine height ($60$ meters),
ground roughness ($0.3$ meters), turbine rotor radius ($20$ meters), and a wind farm of size $2km \times 2km$.\footnote{The implementation of constructing the 12 standard WFLO instances can be found in~\url{https://bit.ly/2YIjTl1}.}

@@

The existing constraint programming approach was
reported to perform worse than the ILP model~\cite{Zhang2014} and so was omitted from our the experiments.
We used state-of-the-art commercial software, Gurobi~\cite{gurobi}, to solve WFL-QCOP, ILP-SS and QP-SS. 
For consistency, we shall refer to WFL-QCOP problem as QP-LS as it is a quadratic program that uses the LS objective.
In addition, we 
used both Fujitsu's Digital Annealer and Gurobi to solve WFL-QUBO: we shall refer to the former as DA-LS, and to the latter
as QUBO-LS. \todo{Arik: Is it necessary to change the names WFL-QUBO -> QUBO-LS, WFL-QCOP -> QP-LS? This makes it confusing for someone who skims over this paragraph.Can't we use the same names (one or the other) through-out?}

@@

For the DA, we used a single configuration that was found to work best across the different instances. 
The DA was run for increasingly large time limits ranging from $0.7$ seconds to $65$ seconds where we no longer observe an improvement in 
performance. The penalty terms $\lambda_1$ and $\lambda_2$ (see~Eq.~\eqref{QUBO}) are selected such that $\lambda_1= \lambda_2$ and sum of the two penalty terms, namely $\lambda_1+\lambda_2$, 
is set to the highest value that the Digital Annealer allows.\footnote{The DA has an upper limit on the size of objective function coefficients.}

@@

We measured the total expected energy of a solution using the SS expression (see Eq.~(\ref{eq:ss})).
%\footnote{The total mean energy is presented in kW.}.
The only method that directly optimizes SS is QP-SS. Hence, for the other methods, we converted solutions
LS-based solution into the SS expression using Eq.~(\ref{eq:ss}) as proposed in Zhang et al.\ \cite{Zhang2014}. Specifically, let $x = (x_1, \ldots, x_k), k=n^2$ be a turbine placement solution obtained by an LS-based method.
Then, the SS energy expression can be written as, 
$$\sum_{i\in \mathcal{N}}\sum_{d \in \mathcal{D}} \frac{1}{3} x_i \Bigg(u_{id,\infty} \Bigg[1-\sqrt{\sum_{j\in \mathcal{U}_{id}}x_j\bigg( 1-\frac{u_{ijd}}{u_{id,\infty}} \bigg) } \Bigg]    \Bigg)^3.$$

Next, we tested the performance of our approach as function of solution time. We 
compared the average performance of all WFLO methods on all 12 WFLO instances. Since energy levels are not directly comparable between the different instances (due to varying number of turbines and wind farm specifications) we used the \emph{mean relative error} (MRE) measure with respect to the best solution obtained per instance.
%Then, the
%time-varying MRE can be averaged across the 12 scenarios and presented over time.
%In what follows, we demonstrate the construction of the MRE measure.

Recall that $\mathcal{W}$ is the set of our 12 standard WFLO instances. 
Let $b_w$ be the best energy solution (in terms of SS) obtained for an instance $w \in \mathcal{W}$ 
across the different methods.
%For example, the best solution (within 3600 seconds) for the second instance $w_2$ that aims at placing $30$ turbines on a $10\times10$ grid with a single wind regime is attained by QUBO-LS and the SS energy expression is $15742$ (kW); thus, we set $b_w = 15742$.
Then, for every point in time $t \in \{0,\ldots, 3600\}$ (measured in seconds), we can compute the relative error of a given approach per instance $w$. Let $\mathcal{A}$ be the set of solution approaches, $\mathcal{A}= \mbox{\emph{QUBO-LS}}, \mbox{\emph{QP-LS}}, \mbox{\emph{DA-LS}}, \mbox{\emph{ILP-LS}}, \mbox{\emph{QP-SS}}\}$. Further, denote by $b_{w,t,a}$ the best solution attained before time $t$ of approach $a$ for instance $w$. Then, the relative error at time $t$ for approach $a$ on instance $w$ is given by
\begin{equation}
  RE(w,t,a) = \frac{b_w - b_{w,t,a}}{b_w}.
\end{equation} Note that the expression is always positive, since we are solving a maximization problem. We can now average over the different instances at time $t$. The average performance of approach $a$ at time $t$, $MRE(t,a)$ can be computed as, \begin{equation} MRE(t,a) = \frac{1}{|\mathcal{W}|}\sum_{ w\in\mathcal{W}} RE(w,t,a).\end{equation}
We present $MRE(t,a)$ for each of the approaches to demonstrate the speed of convergence of the proposed LS-based methods.

The last measure that we introduce is the speed-up factor $S(a_1, a_2, t, w)$. The measure compares the time that it takes
approach $a_1 \in \mathcal{A}$ to
achieve the same energy level as the energy achieved by approach $a_2 \in \mathcal{A}$
within $t$ seconds when solving instance $w$ (energy levels are measured in terms of the sum-of-squares energy expression). 
Formally, $$S(a_1, a_2, t, w) = \frac{\min \big[\{ t' : E_{SS}(a_1, t',w) = E_{SS}(a_2, t, w) \} \cup \{3600\} \big] }{t}, $$
with $E_{SS}(a, t, w)$ being the SS energy value under approach $a$ for instance $w$ at time $t$. The speed-up factor 
is useful to measure how many times faster (or slower) approach $a_1$ solves problem $w$ compared to approach $a_2$.   
For example, $$S(\mbox{\emph{QP-LS}}, \mbox{\emph{DA-LS}}, 10, w_1)$$ is the ratio between the time that it takes QP-LS to 
reach the same quality of solution as DA-LS achieves after $10$ seconds on problem $w_1$. Since
we limit the runtime of the exact methods by $3600$ seconds, $S$ is a lower bound on the true speed-up factor. 
%Therefore, if for the example above the result is $3600/10$, we shall interpret it as a speed-up factor of \emph{at least} $360$.

@@

We observe that the DA-based solution provides
comparable results to state-of-the-art exact methods. For the harder instances (of $36$ wind directions),
the three quadratic LS-based approaches (DA-LS, QUBO-LS, and QP-LS) dominate the others by finding the best (or equal) solutions in 11 out of 12 instances. For easier instances, 
the linear approach (ILP-LS) finds better solutions.
%\todo{@Chris, Jason: do we have a good explanation for this?}. NO
Table~\ref{tab:results2} shows the same type of results as in Table~\ref{tab:results1}, but now after $3600$ seconds. For the exact technique, we can see that the improvement between 100 and 3600 seconds of run-time is marginal. We also observe that the DA at 10 seconds is, therefore, still competitive with the other approaches,
while QP-LS finds the best solutions in 10 out of the 12 scenarios. 
Our quadratic approaches dominate all six hard scenarios (12 wind directions). %\todo{@Chris: I am not sure that the first table contributes anything beyond the second table. JCB: ``We include the second table to demonstrate that only minor improvements are achieved by the exact techniques between 100 seconds of run-time.''}

@@

@@

The results also show that on average 
the ILP-LS method starts with poor solutions, yet ends up improving over QP-SS and QUBO-LS after around 1250 seconds.

To demonstrate the gain in runtime from using the DA, we compare the average speed-up factor between
the QP-LS (the dominant exact method) to a DA-LS solutions after $1$ and $10$ seconds.
For each $w \in \mathcal{W}$ we computed the mean $S(\mbox{\emph{QP-LS}}, \mbox{\emph{DA-LS}}, t, w)$ for $t \in \{1,10\}$. 
%and $S(QP-LS, DA-LS, 10sec, w)$
%for every $w$ 
%and averaged the speed-up factors across the instances. 
The average speed-up factors were found to be $666$ and $100$ for $1$ and $10$ seconds, respectively. \todo{Arik: What about a table -- perhaps in an Appendix? If not, you should at least quote the min and max speed-up factors and comment on the speed-up comparing DA with the other techniques.}
%\footnote{The speed-up factors  were rounded to have 0 values after the floating point.}

These results show that the DA achieves competitive solutions after running for $1$ second.
Note that since for many of the instances QP-LS did not attain
the energy level of DA-LS at time $t$ (1 or 10 seconds),
%we consider the speed-up factor to be a lower bound, meaning
that
the DA-LS speed-up is actually even higher.
%speed-up factor, which we did not compute due to the time limit of $3600$. 
%This is an encouraging result, especially due to the pessimistic computational complexity of WFLO.

@@

\todo{Arik: Should this be Sectiobn 5 rather than 4.3?}

Our result show empirically that the quadratic programming models that we
introduced in Section~\ref{sec:QUBO4WFLO} (WFL-QCOP and WFL-QUBO) are useful in two ways.
First, the two problems are the basis for the QP-LS and QUBO-LS methods solved with Gurobi. The two methods are powerful approaches to solving WFLO. Specifically, QP-LS yields the best performance (after 3600 seconds) for most problem instances
and especially,
for the hard instances with 36 wind directions. In addition, the runtime performance of QP-LS is second
only to the Digital Annealer.  Second, the QUBO model can be solved with recent specialized hardware, Fujitsu's DA, as demonstrated by the DA-LS approach. DA-LS yields the best results in 7 out of 12 
instances, shows competitive performance on the other instances, and outperforms all methods in terms of runtime (a speed-up factor of 3 orders of magnitude when the DA runs for $1$ second).

@@

Finally, though we have not emphasized it here, all the approaches presented, including those from the literature, have the advantage of being represented with a declarative mathematical model. Such a representation makes problem extensions significantly easier and allows formal analysis. While model-based  approaches are common in exact optimization such as mixed-integer programming, they are rare for heuristic techniques such as genetic algorithms and meta-heuristics. Many of the state-of-the-art implementations of heuristic techniques are specialized for the particular problem being solved. A change in the problem definition (e.g., the addition of a new constraint) can, therefore, require substantial software changes. Conversely, model-based techniques can be adapted to a revised problem definition via the declarative addition of the new constraint to the model. Further, amenability to formal analysis means that substantial insight can be gained from mathematical treatments of problem classes, leading, as can be seen in mixed integer programming \cite{Bixby07a}, to substantial improvements in problem solving performance.

%State-of-the-art solvers still make use of advanced data structures but they are specialized for the class of problems expressible in the mathematical formalism rather than the problem at hand. Additionally, the mathematical representation of the model allows formal results to be proved. Model-based approaches can therefore provide

@@

In this work, we presented a novel quadratic programming framework
for solving 
wind farm layout optimization. 
The framework is embodied in two models: constrained (WFL-QCOP) and unconstrained binary
 (WFL-QUBO). The former can be used to solve WFLO using advanced optimization software,
while the latter can be mounted onto quantum-inspired machines that can quickly (approximately) solve WFLO. We used the constrained quadratic formulation to prove 
that the computational complexity of WFLO is $\mathcal{NP}$-hard. Nonetheless,
our empirical evaluation shows that when exact solvers like Gurobi~\cite{gurobi} are
coupled with our WFL-QCOP and WFL-QUBO models they are able to solve small WFLO instances to optimality in less than 3600 seconds.
We also show that when an advanced hardware is used with the QUBO representation, large WFLO instances can be solved quickly with no significant degradation in the total energy compared to exact methods.
This runtime improvement and state-of-the-art performance in terms of the total energy of the resulting solutions makes our quadratic framework useful for designing new wind farms, while varying multiple input parameters.
